---
title: "Handout 9"
author: "Student von Student III"
output:
  pdf_document: default
  html_document: default
bibliography: bib/handout9.bib
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, comment = "")
knit_engines$set(text = function(options) {
  if (options$echo && options$eval) knit_child(text = options$code)
})
```

In this handout, we are going to revisit the Reinhart and Rogoff data. 
We are going to try to uncover, if we can, any causal relationship between 
debt-to-GDP ratios and growth rates.

## Topics and Concepts Covered

- Fitting and interpreting a linear model
- Granger-causality with panel data
- Fixed-effects for giving each observation its own baseline

## R Commands Covered 

- More on conditional plots using `coplot`
- Adding fixed effects in `lm` using `factor`
- Using `summary$coef` to extract coefficients, standard errors, t-values, 
  and p-values from `lm`

# Reinhart and Rogoff: An Analysis

To study sovereign debt and economic growth, the economists Carmen Reinhart 
and Ken Rogoff [@Reinhart.Rogoff2010] have constructed an impressive data 
set that encompasses over 70 countries and 800 years worth of observations 
including variables such as inflation, exchange rates, debt-to-GDP ratios and 
categorical 
variables indicating whether a country is suffering from a debt, inflation 
or banking crisis. While the research is multi-faceted, the aspect that has 
received the most public attention is the claim that once a country's debt-to-GDP 
ratio crosses a threshold of approximately 90%, growth rates plummet.  

This argument is not strictly academic; policy makers in both the US and 
Europe have been arguing over whether the current economic downturn should be 
greeted with more or less government spending, viewpoints I will be referring 
to as "Keynesian" and "austerian."  In a nutshell, the Keynesian argument 
is that governments should spend less during times of economic growth and more 
during times of economic downturn.  Normally, this process is handled by the 
Federal Reserve, through adjusting interest rates (through buying or selling US 
Treasury Bonds). 

At the moment, interest rates are approximately zero.  Since the Federal 
Reserve cannot lower interest rates below zero, the Fed's direct policy 
interventions have been exhausted (QE1 and QE2 are indirect). Government 
spending is the only alternative when interest rates are at zero, since 
private spending is not able to fill the output gap in the short-run. 
The Keynesians argue that the government should assume more debt when interest 
rates are around zero, to fill this output gap.  

The austerian argument leads to the opposite policy prescription.  Austerians 
argue that spending through deficit-financing will lead to the accumulation of 
more debt, increasing the risk of future sovereign default.  This increase in 
risk will raise interest rates in the future.  Since these future expectations 
are factored into today's investment and consumption decisions, increasing debt 
will place an upward pressure on *present* interest rates, slowing down current 
growth.

### Reinhart and Rogoff: The Controversy

As mentioned above, Reinhart and Rogoff provide(d) a key piece of empirical 
evidence to the austerians: countries above the debt-to-GDP threshold of about 
90% face precipitous growth declines.  A recent study, put out by Herndon et al. 
at U. Mass-Amherst, shows that the 90% threshold is a result of poor data 
analysis by Reinhart and Rogoff.  We are going to build off of Herndon's analysis.

Though not required for his handout, I recommend referring to the following articles if you are interested in this data and the debate:

- The original paper positing the 90% threshold: @Reinhart.Rogoff2010
- Herndon et al.'s reply: @Herndon.etal2014 You should focus on the first 10 
  pages or so
- Stephen Colbert 
  [weighs in](http://www.colbertnation.com/the-colbert-report-videos/425748/april-23-2013/austerity-s-spreadsheet-error?xrs=share_copy) on the issue

A few more resources, if you find the topic interesting, are:

- Reinhart and Rogoff 
  [respond](http://www.nytimes.com/2013/04/26/opinion/debt-growth-and-the-austerity-debate.html?pagewanted=all) in the New York Times to the Herndon et al. paper 
- Herndon's co-authors 
  [respond to their response](http://www.nytimes.com/2013/04/30/opinion/debt-and-growth-a-response-to-reinhart-and-rogoff.html?_r=0), also in the New York Times
- An early critique by Irons and Bivens: [@Irons.Bivens2010]
  [(link)](http://www.epi.org/page/-/pdf/BP271.pdf). See especially the section 
  on causality 
- A [blog post](http://www.nextnewdeal.net/rortybomb/researchers-finally-replicated-reinhart-rogoff-and-there-are-serious-problems)
  and [follow up](http://www.nextnewdeal.net/rortybomb/reinhart-rogoff-week-later-why-does-matter) 
  on 'Rortybomb' laying out the issues involved

In the data folder you will find a file `RR.lags.csv`. 
This file contains the variables considered by Herndon, et al. 
We will focus on the following variables:

- `Country`: The country with observation
- `Year`: The year of the observation
- `Infl`}: Inflation rate
- `dRGDP`: Percent difference in real GDP (i.e. inflation-adjusted) from 
   last year to the current year
- `debtgdp`: Debt-to-GDP ratio
- `lag.debtgdp1`, `lag.debtgdp2`, `lag.debtgdp3`: Lagged values of the 
  debt-to-GDP ratio (i.e. if the observation is made in 1990, these are the 
  debt-to-GDP ratios for that country in 1989, 1988, and 1987)
- `dRGDP1`, `dRGDP2`, `dRGDP3`: Lagged values of the percentage difference 
   in `dRGDP`

## Conditional Plots for Causal Analysis

The Reinhart and Rogoff data pose a particular problem: there is absolutely no 
hope of conducting an experiment, yet answering the central macroeconomic 
question of our data involves discerning a causal claim of some type.  There 
are four different claims we can make:

1. Current GDP growth is caused by past debt (the austerian argument)
2. Current debt is  caused by past GDP growth (the Keynesian argument)
3. The two appear to cause each other
4. There is no  causal relationship between GDP and debt 

We are going to start going through these different claims, in order to 
ascertain which is most consistent with the data.  Let's first read in the 
data.

```{r}
RR <- read.csv("data/RR.lags.csv", header = TRUE)
head(RR)
attach(RR)
```

Next, we are going to look at a conditional plot.

```{r}
coplot(dRGDP ~ lag.debtgdp1 | lag.dRGDP1)
```

(You can ignore the line `missing.rows`.  It is telling you that there is 
some missing data).

We are going to clean up the conditional plot a little bit.  Instead of 2 x 3, 
let's make it a single row, with four panels.  Also, let's get rid of the box 
up top, giving you the range of the conditioning variable.  Finally, we can add 
a smooth curve to each panel, showing if there is any relationship between the 
variables in that panel.  We do this with

```{r}
coplot(dRGDP ~ lag.debtgdp1 | lag.dRGDP1, 
       panel = panel.smooth, number = 4, 
       rows = 1, show.given = FALSE)
```

Note the additions added into the call to `coplot`.  These are


- `panel = panel.smooth` adds the smooth curves to each line
- `number = 4` tells how many boxes to produce for the figure
- `rows = 1` tells how many rows to use for the figure
- `show.given = FALSE` tells whether to include the component up top with 
   the range of the conditioning variables.

### Interpretation

The conditional plot allows us to assess the extent to which debt-to-GDP 
ratios in the previous time are predictive of current growth, if previous growth 
is held relatively constant.  Each column of the conditional plot plots 
previous debt-to-GDP versus current growth rates.  The leftmost column 
considers a set of homogenous observations, in the sense that their previous 
growth rate was low.  The rightmost column considers observations for which 
the growth rate was highest.  From this figure there appears, at best, a 
weak relationship between previous debt-to-GDP and current growth, if we adjust 
for the confounding of previous growth.  This provides some suggestive evidence 
that previous debt is not causing a drop in current growth rates.

## Do Past Values of Debt Predict Future Levels of Growth?

Next, we are going to do the same sort of assessment, except using linear models, 
with `lm`.  We are going to see, first, whether past values of growth and 
debt-to-GDP are predictive of current GDP growth.  We will do this through two 
linear models:
```{r}
lm1 <- lm(dRGDP ~ lag.dRGDP1 + lag.debtgdp1)
lm2 <- lm(dRGDP ~ lag.dRGDP1 + lag.dRGDP2 + lag.debtgdp1 + lag.debtgdp2)
```

To summarize these models, we can simply use `summary`. The results for the 
first model are:
```{r}
summary(lm1)
```
and for the second model are:
```{r}
summary(lm2)
```

Let's focus in on `lm2`.  We have created a regression where we want to see 
if *past* values of debt-to-GDP are predictive of *current* levels of growth, 
after accounting for previous values of *previous* levels of growth.  The 
$p$-values on previous debt-to-GDP (`lag.debtgdp1`, `lag.debtgdp2`) are both 
well above 0.05.  Therefore, these effects are not significant.  

We have just shown that debt-to-GDP does not **Granger-cause** growth.  This 
method, named after Clive Granger, strives to discern causal relationships 
from panel data (like when we observe the same countries over multiple years). 
We cannot use the method to claim that one variable causes another, but we can use 
it to say that one does not Granger-cause another. Basically, if a lagged 
independent variable is not predictive of the current levels of an outcome, 
when including previous values of the outcome as a predictor, then the 
independent variable does not Granger-cause the outcome.  If the lagged value of the independent variable is predictive of the current outcome, when including the 
lagged outcome, then we can't say that the independent variable does not 
Granger-cause the outcome.

If that seems unfortunately circuitous, well, that's what happens when we try 
to extract causal claims from observational data in which an experiment is 
impossible.  We can make headway, and reasonable arguments, but the analysis 
will require some logic and hedging.  

## Adding Fixed Effects

Recall that we are looking at panel data, in that we are observing the same 
observation (country) over multiple time periods.  Often, when people analyze 
panel data, they will add *fixed effects* for each observation.  These 
fixed effects can be thought of as giving each country its own intercept - 
we may want to include a term in the model that is simply an indicator variable 
for being that particular country.  For example, let's do the same analysis as 
above, but with fixed effects by country:

```{r}
lm3 <- lm(dRGDP ~ lag.dRGDP1 + lag.debtgdp1 + factor(Country))
summary(lm3)
```

You will notice in the regression table that each country, except one, gets 
its own intercept.  R will automatically drop the first level, which is 
Australia, in this data.  The fixed effect is a means of capturing 
unexplained heterogeneity across countries.  You do not know what, exactly, is 
unique about, say, Denmark, but the Denmark fixed effect of $-0.865$, gives 
all observations from Denmark a baseline specific to Denmark.  Normally, 
when we encounter panel data, we add fixed effects for each observation 
(`Country`), by time period (`Year`), or both.  

In terms of our analysis, note that the $p$-value on previous debt-to-GDP is no 
longer significant at the 0.05 level ($p = 0.0901$).  It appears that, in this 
model as well, previous debt-to-GDP is not Granger causing current growth.

## Precept Problems

In these problems, you are going to continue the analysis from above.

### Question 1

One observation has a value of `dRGDP` greater than 25!  Find this 
observation, and explain why the value is so high.

```{text}

```

## Question 2

Next, revisit the conditional plot from the first part.  This time evaluate 
whether previous GDP has a predictive effect on current debt-to-GDP, after 
accounting for the confounding of previous debt-to-GDP.  
```{r}

```

Taken together, what 
do these two figures suggest about the causal relationship between debt-to-GDP 
and growth?

```{text}

```

## Question 3

Repeat the same set of regressions from the second part, but try to explain 
current debt-to-GDP, instead of growth.  Please try it without fixed effects, 
and then with different fixed effects specifications (country, year, 
country and year).
```{r}

```


## Question 4

Imagine you are at a conference with David Cameron, Angela Merkel, Barack Obama, 
and Shinzo Abe.  (If you don't know who these people are, please look them up.) 
Ben Bernanke is moderating.  Reinhart and Rogoff just presented their data (which, 
in fairness, involves more than we are looking at).  They end up quoting an 
IMF report, which states:

> Much of the empirical work on debt overhangs seeks to identify the 
> 'overhang threshold' beyond which the correlation between debt and growth 
> becomes negative. The results are broadly similar: above a threshold of 
> about 95 percent of G.D.P., a 10 percent increase in the ratio of debt 
> to G.D.P. is identified with a decline in annual growth of about 0.15 
> to 0.20 percent per year. [(link)](http://www.nytimes.com/2013/04/26/opinion/reinhart-and-rogoff-responding-to-our-critics.html)

When they finish, the audience turns and looks to you.  What do you say?

(You should have no doubt, by now, of my opinion on this issue.  I cannot 
claim Absolute Correctness on this issue.  If you disagree, feel free, and 
back it up however you wish---using this data, other citations, whatever works for you.)

```{text}

```

# References

